Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea

### 2022-10-16 analysis/paper/paper.qmd
finish editing methods section

Tip cross-sectional area (TCSA) of stone artefacts has been used as a ballistically significant standard to discriminate different functional classes of projectile armatures such as spearthrower (a.k.a. atlatl) dart tips, arrowheads and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins]. The TCSA metric indicates the part of the tool that cuts the target's hide and relates to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation]. This method was first proposed by @hughes1998getting and and tested by @shea2006origins by comparing archaeological examples to ethnographically collected samples of known use. @shea2009impact also applied this approach to compare projectile points from Africa, the Levant, and Europe, and claimed that projectile weapons first appeared in Africa. A key advantage of the TCSA metric is its convenience: regardless of point type, only maximum width and thickness measurements are needed for calculating the TCSA value (0.5 × maximum width × maximum thickness) [@sisk2011african; @lombard2020tip]. Later, @sisk2011african proposed an alternate measure, tip cross-sectional perimeter (TCSP), for more accurate measure for the force needed to penetrate a target to a lethal depth while TCSA metric is more associated with cutting aspect. However, TCSP has a few disadvantages that limit its applicability to our case study of Korean stemmed points. The force and penetration depth are not only affected by the stone tip, but also by the mass of the shaft, which cannot be known for most archaeological stone-tipped weapons [@lombard2020tip]. @sisk2011 also mentioned that TCSP cannot be applied to backed pieces that are hafted as projectile armatures.

@@

## Applying TCSA

To answer our research questions, What were stemmed points used for? How diverse were their functions? What are the temporal patterns in stemmed point functions? Are these temporal patterns associated with population dynamics or environmental processes? We calculated TCSA values for stemmed points and explored the usage for the stemmed points depending on raw materials, location, and size. We applied radiocarbon dates to the results to examine the temporal patterns. In the discussion, we compared those temporal patterns with demographic and environmental proxies. We used the results of summed probability distributions (SPD) and annual temperature changes from @park2022did which covers our target period and area. Following the TCSA ranges for different weapon-delivery systems that @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths (@tbl-tcsa-ranges), we excluded 12 artifacts from our dataset. Because the TCSA values of those 12 artifacts from Bupyeongri, Hwadaeri, Haga, Sinhwari, Gokcheon, and Sachang sites are greater than 257. Their TCSA range is from 263 to 2622. Therefore, a total of 161 stemmed points from 32 assemblages unearthed from 26 sites are used for the following analysis.

@@

Selective use of raw material is one of the main technological transitions to the Korea Late Paleolithic. Previously used quartzite and vein quartz were continuously selected for the existing core and flake tools and finer grained materials were newly added to assemblages for producing the newly introduced tools [@seong2004quartzite]. We analyze TCSA values per each raw material and examine why types of raw materials were selected for certain types of tool.

### Regional Pattern

We present TCSA values per site that contains more than 4 stemmed points (more than and equal to 5?) based on the evolutionary concept that people choose their tools according to the surrounding environment and available resources. Through the TCSA distribution, we explore which function(s) the stemmed points had performed at the sites.

@@

Different sizes of stone artifacts can represent different functions. Overall size of stone points has been used as a potential proxy to separate different armature types [@thomas1978arrowheads; @sahle2019assessment]. We chose weight as a size indicator for the stemmed points. The weight is a reliable discriminator between tools and a descriptive attribute which can be determined objectively and which serves as a criterion of likeness or dissimilarity in projectile point description [@fenenga1953weights; @shea2006origins]. Excluding 9 points that we were not able to measure or get the records of weight, we calculate TCSA for 152 artifacts.

@@

We add radiocarbon dates to explore the temporal pattern of the function of stemmed points. We excluded the assemblages that have no radiocarbon dates. We arrange 24 assemblages dated from 45ka to 14.8ka periodically, and compute TCSA.

## Two models for weapon-tip types choice

The introduction of new types reflect major implications for social transformation. For example, the bow and arrow enabled individuals to work without hunting teams and provided opportunities for non-elite hunters to produce their own subsistence or pursue individual wealth [@bettinger2013effects; @angelbeck2014faustian; @rorabaugh2015timing]. In evolutionary terms, we assume that human groups carefully selected a specific stone tool based on its functional advantage over other alternatives once they acquired the necessary physiological and cognitive abilities to innovate [@lombard2022tip]. The choice of weapon-tip types, therefore, reflects the socio-environmental circumstances that they had to encounter and overcome. @eren2022plains claim that Clovis points were more variable in shape than Folsom points because Clovis foragers were exposed to largely unfamiliar landscape and employed their points as multifunctional tools that perform a wider range of tasks including cutting and sawing. Folsom points show a narrow range of variation, indicating a specific task. They used TCSA to assess the morphological variance of the points as well as understand the potential functional costs or benefits of one artifact variant relative to another.

Similar to @eren2022plains's approach, we hypothesize two scenarios that explain the regional or/and temporal pattern of TCSA range. If stemmed points represent a narrow TCSA range, people produced tools that performed certain tasks. There could be less surprise in the surrounding environment. On the other hand, a wide TCSA range indicates that stemmed points could be multifunctional tools and people might need to be prepared for unexpected situations such as moving into an unfamiliar landscape or their patch becoming less productive. We expect to observe various TCSA ranges, which reflect morphologically and functionally different stemmed points, across different regions or environments. However, we acknowledge that lithic morphology is highly dependent on the quality of raw materials and their availability.

@@

## Variation by site (Regional pattern)

>>>>

Tip cross-sectional area (TCSA) of stone artefacts has been used as a ballistically relevant standard to probabalistically discriminate different between functional classes of projectile armatures such as spearthrower (a.k.a. atlatl) dart tips, arrowheads and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins]. It is important to note that the TCSA metric alone cannot unambiguously determine artefact function, it only suggests a best-fit ballistic probability for the points if they were hafted as weapon tips. The TCSA metric indicates the part of the tool that cuts the target's hide and relates to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation]. This method was first proposed by @hughes1998getting and and tested by @shea2006origins by comparing archaeological examples to ethnographically collected samples of known use. @shea2009impact also applied this approach to compare projectile points from Africa, the Levant, and Europe, and claimed that projectile weapons first appeared in Africa. A key advantage of the TCSA metric is its convenience: regardless of point type, only maximum width and thickness measurements are needed for calculating the TCSA value (0.5 x maximum width x maximum thickness) [@sisk2011african; @lombard2020tip]. Later, @sisk2011african proposed an alternate measure, tip cross-sectional perimeter (TCSP), for more accurate measure for the force needed to penetrate a target to a lethal depth while TCSA metric is more associated with cutting aspect. However, TCSP has a few disadvantages that limit its applicability to our case study of Korean stemmed points. The force and penetration depth are not only affected by the stone tip, but also by the mass of the shaft, which cannot be known for most archaeological stone-tipped weapons [@lombard2020tip]. @sisk2011 also mentioned that TCSP cannot be applied to backed pieces that are hafted as projectile armatures.

@@

## Investigating patterns in TCSA values

To answer our research questions about the likely uses of stemmed points, we calculated TCSA values for the stemmed points in our sample, and explored the interaction of TCSA values with raw materials, site location, and artefact size. To examine the temporal patterns in TCSA values we investigated radiocarbon ages associated with the points. We then compared those temporal patterns with demographic and environmental proxies based on summed probability distributions (SPD) and annual temperature changes from @park2022did. We followed the TCSA ranges for different weapon-delivery systems that @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths (@tbl-tcsa-ranges). We excluded 12 artifacts from our dataset with TCSA greater than 257 (263-2622) and thus outside of the range of the comparative data. For this study we include TCSA values from a total of 161 stemmed points from 32 assemblages unearthed from 26 sites.

@@

Selective use of raw material is a key characteristic of the technological transition of the Korean Late Paleolithic. Prior to the Late Paleolithic people used quartzite and vein quartz for core and flake tools. Then, finer grained materials were newly added to assemblages for producing the newly introduced tools [@seong2004quartzite]. We analyzed TCSA values for each raw material type to examine the interaction of raw material types and TCSA values.

### Geographical Patterns

To explore geographical patterns in TCSA values, we retained only sites that contains more than four stemmed points. This resulted in a sample of XXX sites and XXX points. We summarised the distribution of TCSA values for each site to identify variation that might be influenced b environmental differences relating to the different locations of the sites.

@@

Different sizes of stone artifacts can constrain or enable different functions. Overall size of stone points has been used as a potential proxy to separate different armature types [@thomas1978arrowheads; @sahle2019assessment]. We chose weight as a proxy for overall size of the stemmed points. Weight is a reliable discriminator between tools of different sizes and a descriptive attribute which can be determined rapidly and objectively and serve as a criterion of likeness or dissimilarity in projectile point description [@fenenga1953weights; @shea2006origins]. Excluding nine points that we were not able to measure or obtain the records of their weight, we explored the relationship between TCSA and weight for 152 artifacts.

@@

We used XXX radiocarbon ages to explore the temporal pattern of the function of stemmed points. We excluded assemblages that have no radiocarbon dates. We arranged 24 assemblages dated from 45ka to 14.8ka to explore variation in TCSA over time.

## Modelling weapon-tip type selection

The process of the introduction of new technologies can vary depending on the environmental and social contexts of the transmission of the techniques for making the new tools, and the new tools themselves can alter the typical contexts of tool use. For example, the introduction of the bow and arrow in American enabled individuals to work without hunting in teams and provided opportunities for non-elite hunters to produce their own subsistence or pursue individual wealth [@bettinger2013effects; @angelbeck2014faustian; @rorabaugh2015timing]. Taking an evolutionary approach, we assume that, given an opportunity to explore alternative technologies, human groups carefully selected a specific stone tool technology based on its advantages over other alternatives, according to their performance in a variety of domains, such as physical and social functions [@lombard2022tip]. Thus, the selection of weapon-tip types is likely to reflect the socio-environmental circumstances that people encountered and managed. For example, @eren2022plains used TCSA to compare the morphological variance of Clovis and Folsom points and claimed that Clovis points were more variable in shape than Folsom points because Clovis foragers were exposed to largely unfamiliar landscape. Clovis points were used as multifunctional tools that performed a wider range of tasks, including cutting and sawing. Folsom points show a narrower range of variation, indicating a they were more likely used for specific tasks.

Inspired by @eren2022plains's approach, we hypothesized two basic scenarios that might explain the regional or/and temporal pattern of TCSA range. If stemmed points represent a narrow TCSA range, then people likely produced tools that performed a small set tasks. This may be due to low levels of uncertainty in the forager's social and physical environments. On the other hand, a wide TCSA range may indicate that stemmed points were multifunctional tools, suggesting that people were responding to unfamiliar situations, such as moving into an unfamiliar landscape or unpredictable changes in patch productivity and travel times \[CITE PAPERS ON ARCHAEOLOGY AND PATCH CHOICE MODEL\]. We expect to observe different TCSA ranges, reflecting morphologically and functionally different stemmed points, across different regions or environments in Korea, and over time.

@@

## Geographical Patterns
